R Notebook

### 2016-07-22 finalwriteup/draft.rmd
finished lit review kerrie comments

Decision trees are one of the simplest and most intuitive machine learning methods. There are several brands of basic decision tree algorithms including ID3, C4.5, CART, and CHAID with CART and C4.5 being the most popular [@Kabra2011]. Ravi Kumar and Ravi [@Kumar2007] recommended the CART algorithm as it is capable of solving both classification and regression problems whereas the remaining decision trees solve classification problems only. Single decision trees tend to overfit the data and provide low predictive accuracy but on the other hand create intuitive rules that a decision maker can follow in real-life scenarios [@Kabra2011][@Putler2012].

To combat the problems of low predictive accuracy and instability, ensemble tree methods were pioneered in the 1990's [@Breiman1996]. Three examples of these methods are boostrap aggregating (bagging), boosting, and Random Forests. Bagging creates multiple trees by sampling a different training data set (*with* replacement) for each tree and then combining tree results [@Breiman1996]. Random forests is an advanced form of bagging and builds on the idea of training multiple trees from the same data by sampling bootstrapped training sets with replacement. However, when creating each tree, a random subset of attributes (variables) is considered at each split. The reduced subset of attributes is resampled for each split in the tree. This allows dominant variables to be suppressed for a fraction of the splits, allowing the algorithm to explore signals in weaker variables. It also prevents the trees from becoming too correlated [@Breiman2001a]. The boosted decision tree approach applies gradient descent theory to a series of decision trees. The trees are limited to a certain depth to maintain simplicity, and each tree models the residuals (or errors) of the preceding tree. The limited depth of each tree prevents overfitting at each stage and the combined result of up to thousands of trees is very powerful [@Elith2008]. The three methods of combining hundreds or thousands of trees turn decision trees into high performing, stable predictors.

Ensemble decision trees have a different foundation to traditional prediction methods such as regression. For this reason, there are some advantages and disadvantages. Decision trees use a progression of binary splits which means the data is first split by one variable at a certain value into two subsets. Then each subset is split by another variable and so on in order to partition the response variable values into groups with similar values. The aim of each split is to achieve the most 'pure' split of the response values in a greedy fashion. If this process is visualised graphically, it is clear that non-linear relationships can be captured.

@@

Similarly to Linear Regression, the predictor variables do not need to be transformed as no assumptions need to be made about the data's statistical distributions [@Louppe2014][@Radenkovic2010]. However, it is often wise to transform skewed distributions to achieve more even segmentation. Other advantages are that ensemble tree methods are faster than SVM's and Neural Networks but perform just as well and even provide insights such as variable importance and variable relationships [@Sealfon2012]. Random forests in particular have the benefit of being unaffected by noise in the data and has the capacity to handle a large number of variables [@Sealfon2012]. Caruana [@Caruana2006] tested Boosted Trees, Random Forests, Neural Networks, SVM's, Logistic Regression and Naive Bayes on 11 binary classification problems and found that Boosted Trees performed best, followed by Random Forests. This demonstrates how ensemble tree methods are capable of competing with high level machine learning algorithms.

@@

A Bayesian Network is a graphical probabilistic model that illustrates the conditional dependencies between variables in a data set. The model is visually represented by a directed acyclic graph (DAG) and is capable of linking the conditional dependency between any variable to another variable. The links are graphically represented by arcs or edges [@Heckerman1998]. This kind of conditional relationship is Bayesian, where the probabilities in one node are conditional upon values in nodes directed to it as well as preceding nodes. Bayesian networks have found success in combining deterministic models with observed data as well as expert knowledge.

Bayesian networks are excellent at representing knowledge from different origins, and then visually communicating results well to decision makers [@Kragt2009]. Therefore Bayesian Networks have become an excellent decision support tool. Network relationships can be machine learned, however this is not widely included as part of the suite of machine learning methods and is usually recommended for modelling complex networks. Drawbacks of Bayesian Networks include their inability to handle continuous numeric variables (all continuous variables must be discretised). This ensures all variable relationships are categorical [@Kragt2009]. Popular applications of Bayesian Networks are modelling uncertainty in natural resource management and modelling complex business network structures such as airports [@Wu2013].

### SVM's and Neural Networks

Neural networks were inspired by biological Neural Networks of the human nervous system and require a lot of training data relative to other methods. They are flexible in the range of relationships they can mimic and are not greatly dependent on the skills of the analyst (Putler & Krider, 2012). A disadvantage that has been mentioned earlier is that the algorithm is a black box. Neural networks can perform high level tasks as a result of deep learning such as hand writing recognition, vehicle control, face recognition, and cancer detection [@Haykin2004].

The simplest type of SVM's calculate the linear boundary between two categories in feature space so that the category of future cases may be determined depending on which side of the boundary they fall. The location of the boundary is determined by the position of the widest clear margin that can be drawn between data points from the two categories, while allowing a certain number of misclassifications (this is usually optimised for the problem). The simple plot below demonstrates this concept and has no misclassifications:

@@

The concept can be adapted to nonlinear boundaries by performing kernel transformations on prior variables as well as regression problems by introducing loss functions [@Auria2008]. Advantages of SVM's include the ability to model nonlinear relationships in data, and that no assumptions about the distributions of the data must be made as the model is non-probabilistic. They are also known to be robust over different samples and perform well in high-dimensional feature space [@Auria2008]. On the other hand, the output is in the form of distance to the boundary (as opposed to probability) and results are not transparent (similar to Neural Networks) [@Caruana2006]. Also, SVM's are known to be very slow to train and therefore often not suitable for industry purposes [@Auria2008]. Because SVM's perform well in high-dimensional space and can form non-linear rules they have been applied to protein classification (medicine) and text and image recognition [@Byun2002].

@@

This section reviewed the advantages and disadvantages of statistical and machine learning techniques that have been successfully applied to business problems. These included Linear Regression, Logistic Regression, Naive Bayes, decision trees and ensemble trees, Bayesian Networks, Neural Networks, and SVM's. Linear/Logistic Regression and Naive Bayes are computationally very fast with a simple concept that make some parametric assumptions about the variable relationships, but work well as baseline models against which to compare complex models. In certain cases these simpler models are significantly outperformed by complex models however it is important to compare both. Neural networks and SVM's are excellent machine learning predictors but are slow black box algorithms and are not appropriate for the effort estimation problem.

Ensemble tree methods are capable of performing as well as Neural Networks and SVM's and make no assumptions about the structure of the data, but provide insights into the model such as variable importance and variable relationships. Bayesian networks are excellent for visually presenting the structure and reasoning behind the predictive model, however are very slow to train and have generally been excluded from the machine learning literature in comparison to the other methods. This may be due to low awareness of learned Bayesian Networks or lower performance.

None of the methods output statistical uncertainty or a confidence interval associated with each prediction. Understanding uncertainty should always be sought in decision making be it from a confidence interval or sensitivity testing. One way to achieve this would be to build hundreds of models using the same method but based on different training divisions of the data. This would produce a range of results for each case, and a statistical confidence interval or classification probability can be calculated for each prediction. Another way for a decision maker to get a feel for uncertainty is to perform sensitivity testing on a prediction. This means the user would run a case through a model several times, each time altering one of the explanatory variable to test the degree of variation in the model predictions. Both strategies would give an idea of the certainty with which a model is predicting a response value or category.

>>>>

Decision trees are one of the simplest and most intuitive machine learning methods. There are several brands of basic decision tree algorithms including ID3, C4.5, CART, and CHAID with CART and C4.5 being the most popular [@Kabra2011]. Ravi Kumar and Ravi [@Kumar2007] recommended the CART algorithm as it is capable of solving both classification and regression problems whereas the remaining decision trees solve classification problems only. Single decision trees create intuitive rules that a decision maker can follow in real-life scenarios but on the other hand, they tend to overfit the data and provide low predictive accuracy [@Kabra2011][@Putler2012]. Ensemble trees were developed to solve this problem and several ensemble techniques are described in this section.

Decision trees have a different theoretical foundation to traditional prediction methods such as regression. They are created by a progression of binary splits of the dataset based on response variable values. Before a split is made, each explanatory variable and each value within the explanatory variables is tested for how 'purely' it partitions the response variable into two subsets. For example, the explanatory variable 'gender' might split a response variable 'hair length' of a class of students into two subsets of students with more similar hair lengths. This similarity is called 'purity' and can be measured using several formulas. An example for regression is:

$SSN - (SSL + SSR)$

$SSN = \sum(y_i - \bar{y})^2$

where SSN is the sum of squared errors of a node (dataset before the split). $y_i$ are the response variable values and $\bar{y}$ is the mean response variable value in the node. SSL and SSR are the sum of squared errors in the left and right datasets that resulted from the split. Other purity formulas exist for categorical response variables such as Gini purity and information gain [@Hastie2009]. After the first split is made on the dataset, each subset is split by another variable (or the same explanatory variable again) and so on resulting in final subsets of data with similar response variable values. If this process is visualised graphically, it is clear that non-linear relationships can be captured.

@@

Once the splitting rules have been learned, a new case is run through the tree from the top. It follows the path provided by which side of the binary split its explanatory variable values fall under and ends up in a final node. The mean response variable value from the initial data that ended up in that node is the predicted response value for the new case. Decision trees are easy to overfit on the training data. This means that if the splitting rules are continued until no more splits can be possibly made, when the test data is run through the tree, the predictions will be too closely aligned to the exact structure of the training dataset. Prediction rules need to generalise trends found in the training data. This can be managed in a single tree by using 'stopping rules' such as a minimum number of data in a final node or a minimum purity increase for a split to be valid (**ref**).

As stated previously, single decision trees also suffer from low predictive accuracy and instability, so to combat these problems, ensemble tree methods were pioneered in the 1990's [@Breiman1996]. Three examples of these methods are boostrap aggregating (bagging), Random Forests, and boosting. Bagging builds multiple trees from different training sample data sets (wich are sampled from the full data set *with* replacement) and then combines tree results [@Breiman1996]. The method of combination is generally averaging numeric predictions or taking the majority rule for categorical variables.

Random forests is an advanced form of bagging and builds on the idea of training multiple trees from the same data by sampling bootstrapped training sets with replacement. However, when creating each tree, a random subset of attributes (variables) is considered at each split. The reduced subset of attributes is resampled for each split in the tree. This allows dominant variables to be suppressed for a fraction of the splits, allowing the algorithm to explore signals in weaker variables. It also prevents the trees from becoming too correlated [@Breiman2001a].

The boosted decision tree approach applies gradient descent theory to a series of decision trees. The trees are limited to a certain depth to maintain simplicity, and each tree models the residuals (or errors) of the preceding tree. By modeling the errors, misclassified cases are weighted higher than correctly classified cases and influence the structure of the tree more [@Hastie2009]. This increased weighting on errors is why the algorithm is called boosting.

This idea can be visualised in a simple scenario consisting of a single explanatory variable (shown on x-axis) and response variable (y-axis). Each tree is limited to one split, and the first tree is shown below, where the vertical line is the splitting point:

![First tree in short progression of boosted trees [@Ihler2012].](images/lit_review/boost_tree1_.png)

The horizontal lines indicate the mean response variable value in each subset after the split. It is the value the first tree would predict for data that falls into each subset. The next tree is built on the residuals of the preceding tree.

![Second tree modeling residuals in short progression of boosted trees [@Ihler2012].](images/lit_review/boost_tree2_.png)

The y-axis represents the distance each point was from the predicted response value (horizontal lines) in the preceding tree. The new split is represented by the vertical line, which creates two subsets of data with more similar residual values. Again, the horizontal lines are mean residuals from the previous tree, which become the predicted residual for each point. The boosting algorithm combines the series of tree predictions. In this case, this is done by adding the predicted residuals from the second tree to the predicted response variable values of the first tree. Visually, the y values of the horizontal lines are added:

![Graphically combined sequence of two boosted trees [@Ihler2012].](images/lit_review/boost_add2.png)

The combination of tree predictions and residual predictions becomes the final predicted response values in boosted trees. The limited depth of each tree prevents overfitting at each stage and the combined result of up to thousands of trees is very powerful [@Elith2008]. The three methods of combining hundreds or thousands of trees turn decision trees into high performing, stable predictors.

The advantages of ensemble tree methods are numerous. Similarly to Linear Regression, the predictor variables in trees do not need to be transformed as no assumptions need to be made about the data's statistical distributions [@Louppe2014][@Radenkovic2010]. However, it is often wise to transform skewed distributions to achieve more even segmentation. Additionally, ensemble tree methods are faster than SVM's and Neural Networks but perform just as well and even provide insights such as variable importance and variable relationships [@Sealfon2012]. Random forests in particular have the benefit of being unaffected by noise in the data and has the capacity to handle a large number of variables [@Sealfon2012]. Caruana [@Caruana2006] tested Boosted Trees, Random Forests, Neural Networks, SVM's, Logistic Regression and Naive Bayes on 11 binary classification problems and found that Boosted Trees performed best, followed by Random Forests. This demonstrates how ensemble tree methods are capable of competing with high level machine learning algorithms.

@@

A Bayesian Network is a graphical probabilistic model that illustrates the conditional dependencies between variables in a data set. The model is visually represented by a directed acyclic graph (DAG) and is capable of linking the conditional dependency between any variable to another variable. The links are graphically represented by arcs or edges [@Heckerman1998]. This kind of conditional relationship is Bayesian, where the probabilities in one node are conditional upon values in nodes directed to it as well as preceding nodes. Refer to the simple example below:

![Simple 4 node Bayesian Network](images/lit_review/BN_1.png)

If each variable, A, B, C, and D are binary categories, there are $2^4$, or 16 combinations of results and each variable is linked to each of the other variables with a conditionally dependency. The probability of each combination could be calculated as per below using the chain rule for probabilities:

$p(ABCD) = p(C|ABD)\cdot p(A|BD) \cdot p(B|D) \cdot p(D)

However, if it is known through domain knowledge or can be reasoned through the data that some variables are conditionally independent from one another, the calculations can be simplified [@Barber2012]. For example, if B and D, A and D, and A and C are conditionally independent, the DAG would be:

![4 node Bayesian Network with conditional independencies](images/lit_review/BN_2.png)

Now, the dependencies that no longer exist can be deleted from the equation:

$p(ABCD) = p(C|BD)\cdot p(A|B) \cdot p(B) \cdot p(D)

In this way, Bayesian Networks reduce computations required to find the probability of an event given other explanatory variable values while still taking into account many conditional dependencies that Naive Bayes cannot [@Barber2012]. Calculating the conditional probability tables for each variable is still much more computationally expensive than Naive Bayes though (as Naive Bayes is the simplest form of a Bayesian Netwoork) and variables with many categories dramatically increases the expense [@Zhang2004].

Network relationships can be learned from the data, however this is not widely included as part of the suite of machine learning methods. The conditional variable dependencies are calculated from the data which in turn can define the graph structure. Some conditional dependencies can be set before the structure is learned [@Barber2012].

Drawbacks of Bayesian Networks include the difficulty for them to process continuous variables. These demand sampling from conditional density models or complex algebra whereas categorical variables require comparitively simple calculations [@Hofmann1996]. It is therefore common to discretise continuous variables [@Kragt2009].

Bayesian networks have found success in combining deterministic models with observed data as well as expert knowledge. They are excellent at combining knowledge from different origins, and then visually communicating results well to decision makers [@Kragt2009]. Therefore Bayesian Networks have become an excellent decision support tool. Another benefit of Bayesian Networks as decision support tools is they allow the user to fix certain variable values, re-run the network and observe changes in probabilities. Thisis a kind of sensitivity testing that gives users a feel for whether the model makes sense and the degree of uncertainty. Popular applications of Bayesian Networks are modelling uncertainty in natural resource management and modelling complex business network structures such as airports [@Wu2013].

### Neural Networks

Algorithmic Neural Networks were inspired by Neural Networks of the human nervous system. Biological Neurons recieve electro-chemical input from the terminals of adjacent neurons into its dendrites. If enough electro-chemical input is recieved, the neuron will fire, which means a signal will be sent from the dendrite end, through the axon to its terminal. Its terminal then sends signals to adjacent dendrites of other neurons and the process propagates. Neurons are able to learn over time by strengthening the connections of adjacent neurons that frequently cause the next one to fire. This means the required firing threshold for the recieving neuron lowers [@Putler2012].

Algorithmic Neural Networks replace neurons with layers of nodes set out in the following structure:

![Layer structure of a Neural Network [@Karpathy2016]](images/lit_review/NN_structure.PNG)

In fact, a wide variety of neural network structures and dynamics exist so a simple popular version is explained here. The first input layer is passive, and each input variable is passed through one input node. These nodes are passive and the variable values are simply passed to each of the nodes in the next layer, called a hidden layer. Neural networks may have any number of hidden layers with any number of nodes as selected by the analyst. Often however, a single hidden layer is adequate. Calibration for predictions is done in the hidden nodes. At these nodes, each input variable is weighted then combined (similar to linear regression) to produce a single value. This is then transformed and assessed by a threshold function to determine if the hidden node should 'fire' to the next nodes. Once the signals and values have propogated to the final layer they are recombined to produce an output, or a prediction [@Putler2012].

The algorithm learns from the data by starting with trial weights for each node. This produces a set of predictions from which errors are calculated. Then optimisation techniques are used to adjust the weights and reduce the error. This process can result in overfitting, so stopping rules such as regularisation strength should be used [@Karpathy2016].

Because of the interwoven network of nodes, Neural Networks can model  complex non-linear relationships. An example of a possible decision boundary in bivariate space is shown below:

![Non linear decision boundary created by a Neural Network[@Karpathy2016]](images/lit_review/NN_bivariate.PNG)

Clearly, a flexible range of relationships can be mimicked by Neural Networks, and if more nodes are added, more flexibility is available [@Putler2012]. A previously mentioned disadvantage is that the algorithm is a black box because the internal structure is too complex for interpretation. They also require a lot of training data relative to other methods. Despite these disadvantages, Neural Networks have found success in high level tasks where the model structure does not need to be understood. These include hand writing recognition, vehicle control, and face recognition [@Haykin2004].

### SVM's

SVM's are non-probabalistic learning algorithms that are typically used for binary classification, although multi-class and regression problems can be solved. It works by geometrically determining the best decision boundary, called a hyperplane, between two classes in multi-dimensional space [@Byun2002] . A hyperplane is simply a plane in more than 3 dimensions, i.e. more than 3 explanatory variables. The simplest type of SVM resembles Logistic Regression because the hyperplane is a linear boundary, and like Logistic Regression, the category of future cases depends on which side of the boundary they fall.

The method for finding the boundary, however, is different to Logistic Regression. The aim is to find the hyperplane with the widest clear margin between data points from the two categories, while allowing a certain number of misclassifications (this is usually optimised for the problem). To do this, the equation for the hyperplane is expressed as follows:

$\boldsymbol{w} \cdot \boldsymbol{x} + \beta_0 = 0$

where $\boldsymbol{w}$ and $\boldsymbol{x}$ are vectors with lengths corresponding to the number of explanatory variables. For two variables, $\boldsymbol{w}$ would be $\begin{pmatrix}\beta_1 \\ \beta_2\end{pmatrix}$ and $\boldsymbol{w}$ would be $\begin{pmatrix}X_1 \\ X_2\end{pmatrix}$ [@Fletcher2009]. Expanding the dot product of these two vectors gives the equation for a line:

$\beta_1 X_1 + \beta_2 X_2 + \beta_0 = 0$

The plane is expressed with vectors because they can be defined such that $\boldsymbol{w}$ is perpendicular to $\boldsymbol{x}$. This aids in optimisation calculations to find the hyperplane with the widest margin, i.e. two hyperplanes with the largest perpendicular distance between them and less than the allowable misclassified points [@Fletcher2009]. The plot below demonstrates this concept and has no misclassifications:

@@

The hyperplanes at the boundaries of the margin are defined by the points closest to the margin. These points are called Support Vectors and during optimisation, these are the only points that matter in defining the final hyperplane position and direction [Auria2008].

The SVM concept can be adapted to nonlinear relationships by performing kernel transformations on the explanatory variables [@Auria2008]. There are several types of kernel transformations that can be applied, but the common aim is to add another dimension to the data points. This can enable a linear hyperplane to be found in multi dimensional space which separates the response variable classes better [@Fletcher2009]. A simple example of this is shown below. The left plot shows two classes that cannot be cleanly separated with a plane. After a radial basis kernel transformation, the binary data in the right hand plot could be separated with a tilted linear plane:

![Data set before and after a radial basis kernel transformation for an SVM model [@Fletcher2009]](images/lit_review/SVM_kernel.PNG)

Advantages of SVM's include the ability to model nonlinear relationships in data, and that no assumptions about the distributions of the data must be made as the model is non-probabilistic. They are also known to be robust over different samples and perform well in high-dimensional feature space [@Auria2008]. On the other hand, the output is in the form of distance to the boundary (as opposed to probability) and results are not transparent (similar to Neural Networks) [@Caruana2006]. Also, SVM's can be very slow to train and therefore not suitable for industry purposes [@Auria2008]. Because SVM's perform well in high-dimensional space and can form non-linear rules they have been applied to protein classification (medicine) and text and image recognition [@Byun2002].

@@

This section reviewed the advantages and disadvantages of statistical and machine learning techniques that have been successfully applied to business problems. These included Linear Regression, Logistic Regression, Naive Bayes, decision trees and ensemble trees, Bayesian Networks, Neural Networks, and SVM's. Linear/Logistic Regression and Naive Bayes are computationally very fast with a simple concept that make some parametric assumptions about the variable relationships, but work well as baseline models against which to compare complex models. In certain cases these simpler models are outperformed by complex models however it is important to compare both.

Ensemble tree methods are capable of performing as well as Neural Networks and SVM's and make no assumptions about the structure of the data, but provide insights into the model such as variable importance and variable relationships. Bayesian networks are excellent for visually presenting the structure and reasoning behind the predictive model, however are very slow to train and have been minimally included in the machine learning literature compared to the other methods. This may be due to low awareness of learned Bayesian Networks or lower performance. Neural Networks and SVM's are excellent machine learning predictors but are slow black box algorithms. Because it is so important to engage decision makers with a model that can explain its results, these two methods are not appropriate for the effort estimation problem.

None of the methods output statistical uncertainty or a confidence interval associated with each prediction. Understanding uncertainty should always be sought in decision making be it from a confidence interval or sensitivity testing. One way to achieve this would be to build hundreds of models using the same method but based on different training divisions of the data. This would produce a range of results for each case, and a statistical confidence interval or classification probability can be calculated for each prediction. Another way for a decision maker to get a feel for uncertainty is to perform sensitivity testing on a prediction. This means the user would run a case through a model several times, each time altering one of the explanatory variable to test the degree of variation in the model predictions. Both strategies would give an idea of the confidence with which a model is outputting response values.

### 2016-07-22 finalwriteup/trial_code_eq.rmd
finished lit review kerrie comments

![Plot of Gaussian probability density distributions of two response classes given x and y explanatory values [@Bulatov2010].](images/lit_review/NB_boundary2.JPG)

>>>>

$\boldsymbol{w} \cdot \boldsymbol{x} + \beta_0 = 0$

where $\boldsymbol{w}$ and $\boldsymbol{x}$ are vectors with lengths corresponding to the number of explanatory variables. For two variables, $\boldsymbol{w}$ would be $w\begin{pmatrix}\beta_1 \\ \beta_2\end{pmatrix}$ and $\boldsymbol{w}$ would be $x\begin{pmatrix}X_1 \\ X_2\end{pmatrix}$. Expanding the dot product of these two vectors gives the equation for a line:

$\beta_1 X_1 + \beta_2 X_2 + \beta_0 = 0$
